---
title: "Introduction to streampress"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to streampress}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

**streampress** is a high-performance binary format for sparse matrices. It provides:

- **Fast I/O** with RLE and entropy coding (rANS, Golomb–Rice, varint)
- **Streaming reads** without loading the full matrix into RAM
- **Column slicing** directly from disk
- **Integrated metadata** (obs/var annotation tables)

This vignette demonstrates the full workflow on synthetic sparse data.

## Generating Synthetic Data

```{r generate}
library(streampress)
library(Matrix)

set.seed(42)

# 5000 rows × 500 columns, ~5% fill
m <- 5000L
n <- 500L
nnz <- as.integer(m * n * 0.05)
rows <- sample.int(m, nnz, replace = TRUE)
cols <- sample.int(n, nnz, replace = TRUE)
vals <- rexp(nnz)                       # non-negative, Poisson-like

mat <- sparseMatrix(i = rows, j = cols, x = vals, dims = c(m, n))
mat <- as(mat, "CsparseMatrix")

cat(sprintf("Matrix: %d × %d, %.2f%% fill, %d non-zeros\n",
            nrow(mat), ncol(mat),
            100 * nnzero(mat) / (nrow(mat) * ncol(mat)),
            nnzero(mat)))
```

## Writing and Reading

```{r write_read}
path <- tempfile(fileext = ".spz")

# Write with default options (v2 chunked sparse)
st_write(mat, path)
cat("Written to:", path, "\n")

# Read back
mat2 <- st_read(path)
cat(sprintf("Read back: %d × %d, %d non-zeros\n",
            nrow(mat2), ncol(mat2), nnzero(mat2)))

# Verify numerical identity
# Verify numerical identity (lossless at auto precision for these values)
stopifnot(all.equal(as.matrix(mat), as.matrix(mat2), tolerance = 1e-6))
cat("Round-trip: PASSED\n")
```

## File Metadata

```{r info}
info <- st_info(path)
print(info)
```

The `st_info()` function returns format version, matrix dimensions, non-zero count, creation time, and per-chunk encoding statistics.

## Column Slicing

A key feature of streampress is efficient **column slicing** — reading a subset of columns without loading the full matrix. Use `st_slice_cols()` for exact column selection:

```{r slice}
# Read columns 101–150 directly from disk (exact, any chunk size)
sub <- st_slice_cols(path, cols = 101L:150L)
cat(sprintf("Sliced columns 101–150: %d × %d\n", nrow(sub), ncol(sub)))

# Same result as indexing the full matrix
ref <- mat[, 101:150, drop = FALSE]
stopifnot(all.equal(as.matrix(ref), as.matrix(sub), tolerance = 1e-6))
cat("Column slice: PASSED\n")
```

## Metadata Tables

streampress files can embed **obs** (row) and **var** (column) annotation tables:

```{r metadata}
obs <- data.frame(
  cell_id   = paste0("cell_", seq_len(nrow(mat))),
  cell_type = sample(c("T", "B", "NK", "Mono"), nrow(mat), replace = TRUE),
  stringsAsFactors = FALSE
)
var <- data.frame(
  gene_id = paste0("gene_", seq_len(ncol(mat))),
  highly_variable = sample(c(TRUE, FALSE), ncol(mat), replace = TRUE),
  stringsAsFactors = FALSE
)

path2 <- tempfile(fileext = ".spz")
st_write(mat, path2, obs = obs, var = var)

# Read back metadata only
obs2 <- st_read_obs(path2)
var2 <- st_read_var(path2)

cat(sprintf("obs table: %d rows × %d cols\n", nrow(obs2), ncol(obs2)))
cat(sprintf("var table: %d rows × %d cols\n", nrow(var2), ncol(var2)))

stopifnot(identical(as.character(obs$cell_id), as.character(obs2$cell_id)))
stopifnot(identical(as.character(var$gene_id), as.character(var2$gene_id)))
cat("Metadata round-trip: PASSED\n")
```

## Dense Matrix I/O

For non-sparse data (e.g., embedding matrices, loadings), streampress offers a dense column-panel format (v3):

```{r dense}
set.seed(1)
dense <- matrix(rnorm(1000 * 32), nrow = 1000, ncol = 32)

path3 <- tempfile(fileext = ".spz")
st_write_dense(dense, path3)

dense2 <- st_read_dense(path3)
stopifnot(all.equal(dense, dense2, tolerance = 1e-6))
cat(sprintf("Dense round-trip: %d × %d PASSED\n", nrow(dense2), ncol(dense2)))
```

## Compression Benchmark

```{r compression, fig.cap="Compression time vs. matrix sparsity."}
library(Matrix)

sizes  <- c(0.01, 0.025, 0.05, 0.10, 0.20)
times  <- numeric(length(sizes))
ratios <- numeric(length(sizes))

for (i in seq_along(sizes)) {
  fill <- sizes[i]
  nnz_i <- as.integer(5000 * 500 * fill)
  m_i <- sparseMatrix(
    i = sample.int(5000, nnz_i, replace = TRUE),
    j = sample.int(500,  nnz_i, replace = TRUE),
    x = rexp(nnz_i),
    dims = c(5000, 500)
  )
  p <- tempfile(fileext = ".spz")
  t0 <- proc.time()["elapsed"]
  st_write(m_i, p)
  times[i] <- (proc.time()["elapsed"] - t0) * 1e3  # ms

  raw_bytes  <- nnz_i * 8 + nnz_i * 4 + (500 + 1) * 4
  file_bytes <- file.info(p)$size
  ratios[i] <- raw_bytes / file_bytes
}

par(mfrow = c(1, 2))
plot(sizes * 100, times,
     xlab = "Fill (%)", ylab = "Write time (ms)",
     type = "b", pch = 19, main = "Write time vs. sparsity")

plot(sizes * 100, ratios,
     xlab = "Fill (%)", ylab = "Compression ratio",
     type = "b", pch = 19, main = "Compression ratio vs. sparsity")
```

Sparser matrices achieve higher compression ratios (e.g., ~`r round(ratios[1], 1)`× at 1% fill) because the run-length and entropy coders exploit longer zero runs. Write time scales roughly linearly with non-zero count.

## Summary

| Operation        | Function          | Key argument        |
|------------------|-------------------|---------------------|
| Write sparse     | `st_write()`      | `obs=`, `var=`      |
| Read sparse      | `st_read()`       | `col_start`, `col_end` |
| File metadata    | `st_info()`       |                     |
| Read obs table   | `st_read_obs()`   |                     |
| Read var table   | `st_read_var()`   |                     |
| Write dense      | `st_write_dense()`|                     |
| Read dense       | `st_read_dense()` |                     |
